---
title: "virolink Quick Start"
format:
  html:
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{virolink Quick Start}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette shows the main user-facing workflow in virolink:

1. read a standardized integration table
2. define host and virus genome objects
3. compose a circos-style integration plot with track objects
4. render a linear integration view for comparison

The example data comes from `inst/extdata/`, and the plots use local
chromosome-size and color tables so the demo does not depend on network access.

## Setup

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(virolink)

extdata_path <- function(file) {
  pkg_path <- system.file("extdata", file, package = "virolink")
  if (!nzchar(pkg_path)) {
    stop("Could not locate extdata file: ", file, call. = FALSE)
  }
  pkg_path
}

chrom_file <- extdata_path("chrom_sizes.txt")
chrom_color_file <- extdata_path("chrom_color.txt")
integration_file <- extdata_path("data.txt")

chrom_colors <- read.delim(chrom_color_file, sep = "\t", stringsAsFactors = FALSE)
host_colors <- stats::setNames(chrom_colors$color, sub("^chr", "", chrom_colors$chr))

integrations <- read_integrations(integration_file)
host <- host_genome(host = NULL, chrom_file = chrom_file, colors = host_colors)
virus <- c(HBV = 3215)
hbv <- hbv_features()
```

## Circos integration plot

The first figure is the main integration view. The track order controls the
stacking order from the outside toward the center.

```{r}
#| label: circos-demo
#| fig-width: 10
#| fig-height: 8
#| message: false
#| warning: false

plot_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  tracks = list(
    track_ideogram(),
    track_sites(
      color = "method",
      size = "support",
      label = "Integration sites"
    ),
    track_virus_genes(
      hbv,
      label = TRUE
    ),
    track_virus_density(
      label = "Virus-position density"
    ),
    track_links(color = "method")
  ),
  visual_ratio = 0.12
)
```

## Linear view

The linear plot gives a simple one-dimensional view of the same integration
records. It is often useful when you want a quick sanity check or a figure that
is easier to read in a report.

```{r}
#| label: linear-demo
#| fig-width: 10
#| fig-height: 4.5
#| message: false
#| warning: false

plot_linear_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  features = hbv
)
```

## Notes

- Use `read_integrations()` for the standardized input format.
- Use `host_genome()` and `virus_genome()` when you want to make the input
  explicit and reusable.
- Build plots with `track_*()` objects and pass them to `plot_integrations()`.
- `plot_oncoprint()` follows the same design philosophy for mutation summaries.
