---
title: "Visualizing Viral Integration with virolink"
format:
  html:
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Visualizing Viral Integration with virolink}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`virolink` organizes visualization around the questions asked of viral
integration data. The main workflow is:

1. standardize integration records
2. inspect their distribution across the host genome
3. connect host breakpoints with positions and features on the virus genome
4. move from cohort-wide patterns to individual loci

The bundled example is deliberately small. All files are loaded from the
installed package, and local chromosome sizes are used so the vignette does
not depend on network access.

## Setup

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

library(virolink)

extdata_path <- function(file) {
  pkg_path <- system.file("extdata", file, package = "virolink")
  if (!nzchar(pkg_path)) {
    stop("Could not locate extdata file: ", file, call. = FALSE)
  }
  pkg_path
}

chrom_file <- extdata_path("chrom_sizes.txt")
chrom_color_file <- extdata_path("chrom_color.txt")
integration_file <- extdata_path("data.txt")

chrom_colors <- read.delim(chrom_color_file, sep = "\t", stringsAsFactors = FALSE)
host_colors <- stats::setNames(chrom_colors$color, sub("^chr", "", chrom_colors$chr))

integrations <- read_integrations(integration_file)
host <- host_genome(host = NULL, chrom_file = chrom_file, colors = host_colors)
virus <- c(HBV = 3215)
hbv <- hbv_features()
```

## Standardize integration records

`read_integrations()` validates the required columns and returns a reusable
integration object. The same workflow accepts an in-memory data frame through
`as_integrations()`.

```{r}
#| label: inspect-data

integrations
```

## Survey the host-genome landscape

Start with the host genome. A binned landscape reveals chromosomes and regions
with recurrent integration events without first committing to a host-virus
link layout.

```{r}
#| label: landscape-demo
#| fig-width: 10
#| fig-height: 4.8
#| message: false
#| warning: false

plot_integration_landscape(
  integrations,
  host = host,
  mode = "density",
  group_by = "method",
  bin_size = 5e7
)
```

Use `mode = "sites"` when supporting-read depth and exact event positions are
more important than regional counts.

## Connect host and virus coordinates

After surveying the host landscape, use a linked view to examine which host
breakpoints connect to which positions on the viral genome.

### Circular linked view

Track order controls stacking from the outside toward the center.

```{r}
#| label: circos-demo
#| fig-width: 10
#| fig-height: 8
#| message: false
#| warning: false

plot_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  tracks = list(
    track_ideogram(),
    track_sites(
      color = "method",
      size = "support",
      label = "Integration sites"
    ),
    track_virus_genes(
      hbv,
      label = TRUE
    ),
    track_virus_density(
      label = "Virus-position density"
    ),
    track_links(color = "method")
  ),
  visual_ratio = 0.12
)
```

### Linear linked view

The linear plot gives a simple one-dimensional view of the same integration
records. It is often useful when you want a quick sanity check or a figure that
is easier to read in a report.

```{r}
#| label: linear-demo
#| fig-width: 10
#| fig-height: 4.5
#| message: false
#| warning: false

plot_linear_integrations(
  integrations = integrations,
  host = host,
  virus = virus,
  features = hbv
)
```

## Planned workflow extensions

The vignette will grow along the same analytical path as new capabilities are
implemented:

- **Breakpoint map:** compare host and virus breakpoint coordinates at scale.
- **Locus context:** inspect genes, transcripts, regulatory annotations, and
  evidence around a selected host breakpoint.
- **Cohort comparison:** summarize integration events across samples, genes,
  host regions, and viral features.
- **Mechanism and clonality:** examine breakpoint signatures and changes across
  tissues or time points when those fields are available.
- **Interactive exploration:** link overview, breakpoint, locus, and tabular
  views for larger datasets.

These sections are roadmap placeholders only. Runnable examples will be added
as each feature becomes part of the public API.

## Workflow notes

- Use `read_integrations()` for the standardized input format.
- Use `host_genome()` and `virus_genome()` when you want to make the input
  explicit and reusable.
- Use `plot_integration_landscape()` before linked views to establish the
  host-genome distribution.
- Build plots with `track_*()` objects and pass them to `plot_integrations()`.
